---
title: "\\fontsize{14pt}{3pt}\\selectfont \\textbf{\\textit{MA678 Midterm Project: Modeling Length of Rehabilitation Stay for Spinal Cord Injury Patients}}"
author: "\\fontsize{12pt}{3pt}\\selectfont Rose Determan"
date: "\\fontsize{12pt}{3pt}\\selectfont December 12, 2021"
fontsize: 11pt
header-includes:
  - \usepackage{titling}
  - \pretitle{\begin{flushleft}}
  - \posttitle{\end{flushleft}}  
  - \preauthor{\begin{flushleft}}
  - \postauthor{\end{flushleft}}  
  - \predate{\begin{flushleft}}
  - \postdate{\end{flushleft}}  
  - \usepackage{wrapfig}
  - \usepackage{caption}
  - \captionsetup{font=scriptsize}
geometry: "left=0.8in,right=0.8in,top=.5in, bottom=.8in"
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    latex_engine: xelatex
    toc: true
    keep_tex: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
pacman::p_load(ggthemes, kableExtra, ggplot2, rstan, rstanarm, ggpubr, DescTools,viridis)
options(warn = -1)
source("clean_data.R")
```

```{r Imports}
#set ggplot theme
theme_set(theme_light())

#Begin import and filter to select data
mdl_data <- f1 %>% dplyr::select(c(UniID, AHDaSyRb, AHDaSyRb_log, AFScorRb,AI2RhADa_log,
                            AInjAge, ANurLvlR_rc, ANurLvlR_rcc,
                            AVertInj, ASpinSrg, AAsscInj,AUMVAdm_rcc,ABMI_st)) %>%
               filter(AInjAge != "0-14y") %>%
               filter(ANurLvlR_rc != "S" & ANurLvlR_rc != "U" & ANurLvlR_rc != "L"& ANurLvlR_rcc != "U") %>%
               na.omit()

w = 6
h= 6/1.62
```

# Abstract
Spinal cord injuries (SCI) are costly and lifelong medical conditions. The first weeks and months after an injury are crucial to recovery, but many variables impact the timeline of patient recovery. Both hospitals and patients can benefit from an accurate prediction of length of stay. In the following analysis, I have modeled the days spent hospitalized in inpatient rehabilitation following an injury using a fixed effect linear model and then a multi-level linear model. Ovarall, although still imperfect, the mixed effect model is more effective and accurate. 

# Introduction
\begin{wrapfigure}{R}{.35\textwidth} 
\begin{center}
    \includegraphics[width=.3\textwidth]{Overview-of-the-Different-Parts-of-the-Vertebral-Column-1-600x543.jpg}  
  \captionsetup[figure]{font=small,labelfont=small}
  \caption{Source: Jones, Olivia. The Vertebral Column, 23 September 2020. https://teachmeanatomy.info/back/bones/vertebral-column/} 
\end{center}
\end{wrapfigure} 

  Spinal cord injuries (SCI) can be an extremely serious condition where the spinal cord is damaged and has a decreased ability to send and receive nerve signals. Symptoms can include paralysis, pain or pressure in head, neck, or back, weakness and/or inability to move any part of the body, and difficulty breathing (Mayo Clinic, 2021).  In the United States, in 2021, it is estimated that there are approximately 54 cases of SCI per 1 million (Jain et al., 2015).  The spinal cord is divided into neurological segments. For analysis, the cervical and thoracic segments have been selected.
  Each named segment roughly corresponds to muscle groups and functions, and damage at a higher level likely indicates more serious impairment. For example, if there is an injury at the C4 level, the individual likely will have weak deltoids (in the shoulder region) and reduced strength and sensation everywhere below this region (Young, 2021).    
  Since 1973 the National Spinal Cord Injury Model Systems have been collecting data in a database. The database is well managed and extensively documented. The data come from 29 facilities across the United States. The database includes information on 32,159 individuals from 1972 to 2016 and includes a variety of fields which include, for example, injury year, age at injury, sex, use of mechanical ventilation, functional independence scores, ASIA motor index scores, and ASIA sensory scores.  This report focus on the modeling the number of days that a patient is hospitalized in inpatient rehabilitation. This is the time after a patient has been treated medically/surgically and before they are discharged from care. This information is relevant to hospitals, so they are able to plan treatment for patients and estimate the number of patients they can care for in a given time. This information is also relevant to patients and their families, so they can also plan the next steps in the patient's recovery.   

# Methods
```{r Variable Definitions, include=TRUE}
vars <- read.csv("variables.csv")
kable(vars, "latex") %>%
   column_spec(1, width = "12em") %>%
   column_spec(2, width = "7em") %>%
   column_spec(3, width = "27em") %>%
   column_spec(4, width = "13em") %>%
   kable_styling(font_size = 7.5)
```
First, I selected relevant variables which are listed in the table. Although the dataset was well organized and documented, I removed missing and unknown values and completed some centering, scaling, and recoding of variables. Any modifications can be found in the ```clean_data.R``` file, and brief notes are listed in the table. I selected cervical and thoracic injuries, since there are 4582 and 2510 cervical and thoracic injuries in the complete data set and only 607 lumbar injuries and 9 sacral injuries listed. In addition, individuals 0-14y were excluded from the analysis due to a high rate of missing data. After removing the missing data and the excluded variables, the dataset includes 5168 individuals with injuries from 2006 to 2016.
Next, I fit a fixed effect model and random effect model to predict log(days hospitalized in inpatient rehab). The random effect model includes a random intercept based on neurologic level of injury and a random slope for the log(days from injury to inpatient rehab admission) which is also based on neurologic level of injury.   
```AHDaSyRb_log ~ AI2RhADa_log + ABMI_st +  AInjAge + AFScorRb + AUMVAdm_rcc + AVertInj+ ASpinSrg+ AAsscInj +  ANurLvlR_rcc```     

```Random effect: AHDaSyRb_log ~  AI2RhADa_log + ABMI_st +  AInjAge + AFScorRb + AUMVAdm_rcc + AVertInj+ ASpinSrg+ AAsscInj + (1 + AI2RhADa_log | ANurLvlR_rcc)```  

I anticipate that the level of injury will impact the "baseline" of days hospitalized. I also anticipate that different levels of injury would also have a varied impact on the influence of log(days) from injury to rehab admit on the log(days) hospitalized in inpatient rehab. The models were evaluated using posterior predictive checks and leave-one-out validation (loo). 


# Results
### Exploratory Data Analysis
```{r Density Figure}
ggplot(mdl_data)+
   geom_density(mapping = aes(x = AHDaSyRb_log, group = ANurLvlR_rcc),alpha = 0.25, color = "antiquewhite4")+
   geom_density(mapping = aes(x = AHDaSyRb_log),color = "#33638dff", size = 1.05)+
   labs(title = "Density of log(Days) Hospitalized: Inpatient Rehabilitation")+
   ylab("Density") + xlab("log(Days Hospitalized in Rehabilitation)")
ggsave("density.png", width = w, height = h)
```
\begin{wrapfigure}{R}{.55\textwidth} 
\begin{center}
    \includegraphics[width=.5\textwidth]{density.png}  
  \captionsetup[figure]{font=small,labelfont=small}
  \caption{Density of log(days) hospitalized in inpatient  acute and subacute rehabilitation unit. The blue line is the density of all selected data which excludes individuals who were never admitted either due to recovery or death. Each gray line represents a distribution of log days for a given neurologic level of injury.} 
\end{center}
\end{wrapfigure} 
The mean log(days) hospitalized in inpatient rehab varies based on level of neurologic injury. I expected high level injuries to have longer hospital stays on average, and this was largely true The average stay was 52 ± 1 days (95% confidence interval of the mean). Nearly 23% of individuals in the dataset had a C04 level injury and an average rehab stay of 62 ± 2 days (95% CI), which was also the longest stay of all injury levels. 
Each selected variable has some relationship with the response variable. Examining the plot of comparisons, a longer time from injury to rehab admit corresponds to a longer stay in rehab, and a higher independence score corresponds to a shorter stay in rehab. Both of these relationships make logical sense. Based on the boxplots, individuals who have additional injures and surgeries have a longer stay in rehab, and individuals who have a need for ventilation also have longer stays, and again these relationships make logical sense when thinking about the impact of an additional injury, for example. 

```{r eda fig }
ts <- 7
inj_hsp <- ggplot(mdl_data, aes(y= AHDaSyRb_log, x = AI2RhADa_log)) +
           geom_point(alpha = 0.1) + ylab("log(Days)") + xlab("log(Days) Injury to Rehab Admit")+
           theme(axis.title=element_text(size=ts),
                 axis.text=element_text(size=ts))

bmi <- ggplot(mdl_data, aes(y= AHDaSyRb_log, x = ABMI_st)) +
           geom_point(alpha = 0.1) + ylab("log(Days)") + xlab("BMI")+
           theme(axis.title=element_text(size=ts),
                 axis.text=element_text(size=ts))

f_scr <- ggplot(mdl_data, aes(y= AHDaSyRb_log, x = AFScorRb)) +
           geom_point(alpha = 0.1) + ylab("log(Days)") + xlab("Independence Score")+
           theme(axis.title=element_text(size=ts),
                 axis.text=element_text(size=ts))

age <- ggplot(mdl_data, aes(y= AHDaSyRb_log, x = AInjAge)) +
           geom_boxplot(notch = TRUE) + ylab("log(Days)") +  xlab("Age (years)")+
           theme(axis.title=element_text(size=ts),
                 axis.text=element_text(size=ts))

vert <- ggplot(mdl_data, aes(y= AHDaSyRb_log, x = AVertInj)) +
           geom_boxplot(notch = TRUE) + ylab("log(Days)") + xlab("Vertebral Injury")+
           theme(axis.title=element_text(size=ts),
                 axis.text=element_text(size=ts))

spn <- ggplot(mdl_data, aes(y= AHDaSyRb_log, x = ASpinSrg)) +
           geom_boxplot(notch = TRUE) + ylab("log(Days)") + xlab("Spinal Surgery")+
           theme(axis.title=element_text(size=ts),
                 axis.text=element_text(size=ts))

assc <- ggplot(mdl_data, aes(y= AHDaSyRb_log, x = AAsscInj)) +
           geom_boxplot(notch = TRUE) + ylab("log(Days)") + xlab("Associated Injury")+
           theme(axis.title=element_text(size=ts),
                 axis.text=element_text(size=ts))


vent <- ggplot(mdl_data, aes(y= AHDaSyRb_log, x = AUMVAdm_rcc)) +
           geom_boxplot(notch = TRUE) + ylab("log(Days) Inpatient Rehab") + xlab("Mechhanical Ventilation")+
           theme(axis.title=element_text(size=ts),
                 axis.text=element_text(size=ts))


eda_fig <- ggarrange(inj_hsp, bmi, f_scr, age, vert, spn, assc, vent, ncol = 4, nrow = 2, align = 'v')
eda_fig
ggsave("eda_fig.png", eda_fig, width = w+2,  height = h, units = "in")
```

\begin{figure}{}
  \centering
    \includegraphics[width=\linewidth]{eda_fig.png}
  \caption{Selected variables compared to log(days) hospitalized in inpatient rehab. These variables were selected for inclusion in the models. Continuous variables are plotted in a scatterplot. BMI has been centered at zero and scaled to have a standard deviation of 1. Categorical or binary variables are displayed as boxpots. }
\end{figure}

Slopes vary for the levels of neurologic injury when comparing the log(days) from injury to rehab admission to log(days) spent hospitalized in inpatient rehab. This justifies the use of a model that includes random slopes for log(days) from injury to rehab admission. As shown in Figure 4, it seems as though higher injury levels have a steeper slope when compared to lower injury levels, which have a slopes closer to zero. 
```{r randomslopefigure}
ggplot(data = mdl_data)+
  geom_point(mapping = aes(x = AI2RhADa_log, y = AHDaSyRb_log), alpha = 0.25, color = "gray" )+
  geom_abline(slope = 1, intercept = 0, color = "gray")+
  geom_smooth(aes(x = AI2RhADa_log, y = AHDaSyRb_log, color = ANurLvlR_rcc), method = "lm",se = FALSE, size = 0.85)+
  xlab("Injury to Rehab Admission log(days)")+  ylab("Inpatient Rehab log(days)")+labs(title = "Injury to Rehab Admit Days vs Days in Inpatient Rehab")+
  theme(legend.position = "bottom")+
  guides(color = guide_legend(override.aes = list(size = 1.25),
                              nrow = 2, byrow = TRUE,
                              title = "Level"))+
  scale_color_viridis(discrete = TRUE,option = "D")


ggsave("days.png", width = w+2,  height = h+2, units = "in")
```

\begin{wrapfigure}{L}{.55\textwidth} 
\begin{center}
    \includegraphics[width=.5\textwidth]{days.png}  
  \captionsetup[figure]{font=small,labelfont=small}
  \caption{log(Days) from injury to rehab admission compared to log(days) spent in inpatient rehab. The lines are colored by the neurologic level of injury. Darker purple corresponds to higher on the spine, and yellow coresponds to lower on the spine. As the injury level decreases, it appears that the slope also decreases. } 
\end{center}
\end{wrapfigure} 
  
  

## Models

```{r Fixed Model}
f_mdl <- stan_glm(formula = AHDaSyRb_log ~ AI2RhADa_log + ABMI_st +  AInjAge + AFScorRb + AUMVAdm_rcc + AVertInj+ ASpinSrg+ AAsscInj +  ANurLvlR_rcc, data =  mdl_data, refresh = 0, iter = 6000)
saveRDS(f_mdl, "f_mdl.rds")

f_mdl <- readRDS("C:/Users/RoseDeterman/2021_Fall/MA678/MA678-project-SCI/f_mdl.rds")

fix_pp <- pp_check(f_mdl,nreps = 100)

params <- names(f_mdl$coefficients)
fix_sp <- stan_plot(f_mdl, point_est = "mean", show_density = TRUE,ci_level = 0.95,outer_level = 0.995,pars = params[2:12], fill_color = "White", outline_color = "#e41a1c")
```

```{r Random Model}
r_mdl <-stan_lmer(formula = AHDaSyRb_log ~  AI2RhADa_log + ABMI_st +  AInjAge + AFScorRb + AUMVAdm_rcc + AVertInj+ ASpinSrg+ AAsscInj + (1 + AI2RhADa_log | ANurLvlR_rcc) ,data = mdl_data, iter= 6000, chains = 5, control = list(adapt_delta = 0.999999))
saveRDS(r_mdl, "r_mdl.rds")
r_mdl <- readRDS("C:/Users/RoseDeterman/2021_Fall/MA678/MA678-project-SCI/r_mdl.rds")
ran_pp <- pp_check(r_mdl,nreps = 100)

params <- names(r_mdl$coefficients)
r_sp <- stan_plot(r_mdl, point_est = "mean", show_density = TRUE,ci_level = 0.95,outer_level = 0.995,pars = params[2:12], fill_color = "White", outline_color = "#377eb8")
```


```{r stan plots}
param_plt <- ggarrange(fix_sp, r_sp, ncol = 1, common.legend = TRUE, labels = c("Parameter Estimates: Fixed Effect Model","Parameter Estimates: Random Effect Model"),
  hjust = -0.7,
  vjust = 1.75,
  font.label = list(size = 10, color = "black", face = "bold"))
ggsave("param_plot.png", param_plt, width = w+1, height = h+1)
param_plt
```
\begin{figure}{}
  \centering
    \includegraphics[width=\linewidth]{param_plt.png}
  \caption{Each plot shows the distribution of the estimated coefficeints for the fixed effect model (red) and the random effect model (blue). The estimates point estimates are similar to eachother, but the distributions are slighly different. Interestingly, those in the 75+ age category have a negative coefficient which indicates fewer days hospitalized in inpatient rehab. }
\end{figure}


\begin{wrapfigure}{r}}{0.55\textwidth}
  \centering
    \includegraphics[width=0.55\linewidth]{param_plot.png}
  \caption{Each plot shows the parameter estimate and density for the nine parameters that are fit in both models. The parameter estimates are largely similar for both models. Interestingly, the estimate for those 75 and older is negative, so individuals in this group spend less time in rehabilitation than someone who is 15-29 years old. It also seems that spinal surgery and associated injuries do not play a strong role in the length of time in rehab.}
\end{wrapfigure}

As mentioned, I fit two models to examine the differences between the fixed effect model and the random effect model. For both I used bayesian methods for the regression. Please refer to the appendix for detailed outputs of both models and model validation. 
A leave one out cross validation that compares the fixed effect model and the random effect model shows that the random effect model is a better fit based on an ```ELPD``` value that is 20 points higher than the fixed effect model. The ELPD is the theoretical expected log point-wise predictive density of the model. The ```p_loo``` value returned from the validation indicates the number of "effective parameters." In the random effect model, approximately 85% (44/52) of the parameters are effective while in the fixed effect model, the ```p_loo``` value is larger than the number of parameters that were estimated which indicates weak predictive power of the model.  

\begin{wrapfigure}{}{0.55\textwidth}
\begin{center}
    \includegraphics[width=.5\textwidth]{cat.png}
  \captionsetup[figure]{font=small,labelfont=small}
  \caption{The caterpillar plots show the point estimate of the random intercepts (top) and slopes (bottom) and the 95% credible interval. The x-axis lists the level of neurologic injury from highest level to lowest when read left to right. The color represents the "rank" of the intercept or slope. When compared to the other slopes is the lowest (rank 1) or highest (rank 20) The dark horizontal line shows the average of the coefficients.}
\end{center}
\end{wrapfigure}

Looking at the caterpillar plots, we can see the random slope and intercept estimates. In general, injuries closer in level to each other have more similar coefficient estimates. It is also interesting to note that if a level has an above average intercept, it also often has a below average intercept. The most prominent of this pattern is for injuries at the T03 level, since it has the largest intercept estimate and the smallest slope estimate. 

```{r caterpillar code}
# Obtain school-level varying intercept a_j
# draws for overall mean
mu_a_sims <- as.matrix(r_mdl,
                       pars = "(Intercept)")
# draws for 73 schools' school-level error
u_sims <- as.matrix(r_mdl,
                    regex_pars = "b\\[\\(Intercept\\) ANurLvlR_rcc\\:")
# draws for 73 schools' varying intercepts
a_sims <- as.numeric(mu_a_sims) + u_sims

# Obtain sigma_y and sigma_alpha^2
# draws for sigma_y
s_y_sims <- as.matrix(r_mdl,
                       pars = "sigma")
# draws for sigma_alpha^2
s__alpha_sims <- as.matrix(r_mdl,
                       pars = "Sigma[ANurLvlR_rcc:(Intercept),(Intercept)]")

# Compute mean, SD, median, and 95% credible interval of varying intercepts

# Posterior mean and SD of each alpha
a_mean <- apply(X = a_sims,     # posterior mean
                MARGIN = 2,
                FUN = mean)
a_sd <- apply(X = a_sims,       # posterior SD
              MARGIN = 2,
              FUN = sd)

# Posterior median and 95% credible interval
a_quant <- apply(X = a_sims,
                 MARGIN = 2,
                 FUN = quantile,
                 probs = c(0.025, 0.50, 0.975))
a_quant <- data.frame(t(a_quant))
names(a_quant) <- c("Q2.5", "Q50", "Q97.5")

# Combine summary statistics of posterior simulation draws
a_df <- data.frame(a_mean, a_sd, a_quant,names(a_mean))
a_df$Level <-   str_extract(a_df$names.a_mean.,"[TC]\\d\\d")
round(head(a_df[,1:3]), 2)

# Sort dataframe containing an estimated alpha's mean and sd for every school
a_df <- a_df[order(a_df$a_mean), ]
a_df$rank <- c(1 : dim(a_df)[1])  # a vector of school rank

# Plot school-level alphas's posterior mean and 95% credible interval
c_int <-ggplot(data = a_df,
       aes(x = Level,
           y = a_mean,)) +
  geom_pointrange(aes(ymin = Q2.5,
                      ymax = Q97.5,
                      color = rank )) +
  geom_hline(yintercept = mean(a_df$a_mean),
             size = 0.5,
             col = "#33638dff") +
  scale_y_continuous(expression(paste("varying intercept, ", alpha[j])))+
  scale_color_viridis_c()


# Obtain school-level varying intercept a_j
# draws for overall mean
mu_a_sims <- as.matrix(r_mdl,
                       pars = "AI2RhADa_log")
#draws for 73 schools' school-level error
u_sims <- as.matrix(r_mdl,
                    regex_pars = "b\\[AI2RhADa_log ANurLvlR_rcc\\:")
# # draws for 73 schools' varying intercepts
 a_sims <- as.numeric(mu_a_sims) + u_sims

# Obtain sigma_y and sigma_alpha^2
# draws for sigma_y
s_y_sims <- as.matrix(r_mdl,
                       pars = "sigma")
# draws for sigma_alpha^2
s__alpha_sims <- as.matrix(r_mdl,
                       pars = "Sigma[ANurLvlR_rcc:AI2RhADa_log,AI2RhADa_log]")

# Compute mean, SD, median, and 95% credible interval of varying intercepts

# Posterior mean and SD of each alpha
a_mean <- apply(X = a_sims,     # posterior mean
                MARGIN = 2,
                FUN = mean)
a_sd <- apply(X = a_sims,       # posterior SD
              MARGIN = 2,
              FUN = sd)

# Posterior median and 95% credible interval
a_quant <- apply(X = a_sims,
                 MARGIN = 2,
                 FUN = quantile,
                 probs = c(0.025, 0.50, 0.975))
a_quant <- data.frame(t(a_quant))
names(a_quant) <- c("Q2.5", "Q50", "Q97.5")
# Combine summary statistics of posterior simulation draws
a_df <- data.frame(a_mean, a_sd, a_quant,names(a_mean))
a_df$Level <-   str_extract(a_df$names.a_mean.,"[TC]\\d\\d")
round(head(a_df[,1:3]), 2)

#Sort dataframe containing an estimated alpha's mean and sd for every school
a_df <- a_df[order(a_df$a_mean), ]
a_df$rank <- c(1 : dim(a_df)[1])  # a vector of school rank

# Plot school-level alphas's posterior mean and 95% credible interval
c_slp <- ggplot(data = a_df,
       aes(x = Level,
           y = a_mean,)) +
  geom_pointrange(aes(ymin = Q2.5,
                      ymax = Q97.5,
                      color = rank )) +
  geom_hline(yintercept = mean(a_df$a_mean),
             size = 0.5,
             col = "#33638dff") +
  scale_y_continuous(expression(paste("varying slope, ", beta[j])))+
  scale_color_viridis_c()
```

```{r caterpillar plot}
cat <- ggarrange(c_int, c_slp, nrow = 2, common.legend = TRUE, legend = "bottom")
ggsave("cat.png", plot = cat,  width = w+1,  height = h+3, units = "in")
```


```{r}
pred <- data.frame(AFScorRb = rep(x = 20, times = 4),
                   AI2RhADa_log = rep(x = 2.639057, times = 4),
                   ABMI_st= rep(x = -0.1589322, times = 4),
                   AInjAge = rep(x = "15-29y", times = 4),
                   AVertInj = rep(x = "1", times = 4),
                   AAsscInj = rep(x = "0", times = 4),
                   ASpinSrg = rep(x = "1", times = 4),
                   AUMVAdm_rcc = rep(x ="0", times = 4),
                   ANurLvlR_rcc = c("C04", "C05", "T10", "T12"))


pred_res <- c()
for (x in c(1:4)){
  r_pred <- posterior_predict(r_mdl, pred[x,], draws = 15000)
  pred_res <- rbind(pred_res, c(pred[x,9],round(mean(r_pred),2), round(sd(r_pred),2)))
}
pred_res <- data.frame(pred_res)
names(pred_res) <- c("Injury Level", "Mean log(days)", "Std.Dev. log(days)")
pred_res
kable(pred_res, "latex") %>%
  kable_styling(font_size = 7.5)
```
I selected the median value for numeric inputs and the most common input for categorical variables, and used posterior predication with 15,000 draws to estimate the predicted log(days) spent hospitalized in inpatient rehabilitation. I also selected the two most common cervical levels injuries and the two most common thoracic levels of injuries. With these simple examples, it is interesting that lower levels of injuries have a longer expected stay in rehab. 

# Discussion
In the random effect model, there were several elements that were surprising and should be investigated further.
It was surprising to that patients with lower levels of injury (less of their body is impacted) were predicted to have longer stays. Maybe this is because they have more body functions, so there are more areas where they can improve and progress, while individuals with high levels of injuries are more limited. It should also be noted that these estimations also have large standard deviations which indicate a larger degree of uncertainty.  
Additionally, I was not expecting that individuals in the 75+ age group would have a shorter duration of rehab hospitalization. Perhaps this is due to a lack of data for this population. SCIs are most common in young people, and in the dataset used for the models less than 5% of patients were in the 75+ group.


# Works Cited
Chen, Yuying. National Spinal Cord Injury Model Systems Database. Version 2016ARPublic. Birmingham, AL: National Spinal Cord Injury Statistical Center [distributor]. https://doi.org/10.17605/OSF.IO/NP24C. Accessed 28 October 2021.

Jain, N. B., Ayers, G. D., Peterson, E. N., Harris, M. B., Morse, L., O’Connor, K. C., & Garshick, E. (2015). Traumatic Spinal Cord Injury in the United States, 1993–2012. JAMA, 313(22), 2236–2243. https://doi.org/10.1001/jama.2015.6250

Spinal cord injury—Symptoms and causes. (October 2, 2021). Mayo Clinic. Retrieved November 19, 2021, from https://www.mayoclinic.org/diseases-conditions/spinal-cord-injury/symptoms-causes/syc-20377890

Young, Wise (2021). Spinal Cord Injury Levels & Classification. Travis Roy Foundation. Retrieved November 29, 2021 from https://www.travisroyfoundation.org/sci/resources/spinal-cord-injury-levels-classification/

# Appendix:
## Mean and CI of rehab stays
```{r, include = TRUE}
mdl_data %>% group_by(ANurLvlR_rcc) %>%
  summarise(days = mean(AHDaSyRb),
            days_ci = MeanCI(AHDaSyRb)[1]- MeanCI(AHDaSyRb)[2],
            count = n()
            )
```


## Summary of fixed effect model
```{r, include = TRUE}
summary(f_mdl)
```


## Summary of mixed effect model
```{r, include=TRUE}
summary(r_mdl)
```

## Random Effect Residual Plot
```{r, include = TRUE}
ggplot() +
  geom_point(mapping = aes(x = r_mdl$fitted.values, y = r_mdl$residuals), alpha = 0.2)+
  geom_hline(yintercept = 0)
```

## PP Checks
```{r pp_checks}
pp_check_plt <- ggarrange(fix_pp, ran_pp, ncol = 1, common.legend = TRUE, labels = c("PP Check: Fixed Effect Model","PP Check: Random Effect Model" ))
ggsave("pp_plot.png", pp_check_plt, width = w, height = h)
```
\begin{figure}{}
  \centering
    \includegraphics[width=\linewidth]{pp_plot.png}
  \caption{Each plot shows the distribution of the response variable log(days) and the distribution of the 100 posterior predicted responses. Both models seem to approximate the distribution well. }
\end{figure}


## Results/Output of LOO
```{r Loo, include = TRUE}
loo_f <- loo(f_mdl)
loo_r <- loo(r_mdl)
loo_compare(loo_f,loo_r )
loo_r
loo_f
```

